Community-Conditioned Language Models

### 2021-10-08 paper/paper1.tex
text

\subsection{Comparing embeddings: method}
% Explain the problem (idea: comparing cosine similarities --> too much data, again)

Let us call \(L\) the matrix of (normed) linguistic embeddings and \(S\) the matrix
of (normed) social embeddings. Thus \(L_i\) and \(S_i\) are the linguistic and
social embeddings for community \(i\).

To compare embeddings, one might naively think that it suffices to
calculate the distance between two embeddings index-wise, which is
equal to the Frobenius distance between \(L\) and \(S\):

@@

The problem with this approach is that even if several dimensions of
\(L\) and \(S\) are capture similar information, they will not
coincide in the \emph{representation}. That is, applying a simple
rotation (orthogonal transformation) on either matrix widely changes
the correlation metric. To make the metric resilient to such effects,
we can compute the minimum distance between \(L_i\) and \(S_i\), for
any orthogonal matrix \(\Omega\) applied to \(L\):

@@

The problem of computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition (SVD) \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

@@

To get a more precise sense of what value of \(d(L,S)\) indicates a significant correlation,
We have experimentally computed \(d(L,S)\) for ten random  ``embeddings'' 
with the same shape as the CCLM embeddings.
We observed a mean of \(μ_d=431.39\) and a (Bessel's-corrected) 
standard deviation \(s_d=2.90\) in their distance from the social embedding, \(S\). % luckily it seems that sample stdev is often written with an $s$ to differentiate from population stdev

Thus if \(d(L,S)\) is below the mean by several standard deviations, we can safely
assume that there is statistically significant correlation between
\(L\) and \(S\).  This is the case: we observe a difference between 61 and 68 standard deviations
(\cref{tab:embedding-correlations}). Further, it is clear that the CCLM embeddings
capture some of the same information as the social-network embeddings.
We also give a sense of \emph{how} the correlation is manifested by
analysis of the two principal components of correlation in the
linguistic embeddings, \(U_0\) and \(U_1\). To do so we plot the
projection of each embedding along their first two principle components
which, together with the corresponding singular values, gives an idea
of how much and in what way they differ (see~\cref{fig:pca-aligned}).

>>>>

\subsection{Comparing embeddings: anecdotal evidence}
\label{sec:storytelling}
TODO: BILL

\subsection{Comparing embeddings: statistical evidence}

The anecdotal evidence suggests that for some communities, proximity
in linguistic patterns do match social proximity… and for others it
does not. Unfortunately, it seems hard, if not impossible, to draw a
general conclusion from this way. In this section, we propose a
systematic approach thanks to which we can quantify the correlation
between social proximity and linguistic proximity, and measure its
statistical significance.

Let us call \(L\) the matrix of (normed) linguistic embeddings and
\(S\) the matrix of (normed) social embeddings. Thus \(L_i\) and
\(S_i\) are the linguistic and social embeddings for community \(i\).
%
A straightforward, but ultimately flawed, generalization of the
methodology used to collect data in \cref{sec:storytelling} is to
consider each pair of communities \((i,j)\), and compute the
correlation between the cosine similarities of both embeddings. That
is, compute the Pearson correlation factor of the data set:

\[\C = {(x = L_i · L_j, y = S_i · S_j) \text {~for~} i,j ∈ [1,512]\}\]

One could then use a standard statistical test to verify to obtain a
p-value for the correlation [TODO: what do we get?].  The flaw of
this methodology is that the number of pairs grows with the square
of the number of communities. With 510 communities, we have 129795
pairs, and standard statistical tests on Pearson correlation will
assure that we have a statistically significant correlation in all
but the weakest correlation. A further flaw is that the distribution of data points
of \(C\) are \emph{not} independent. Far from it, the are generated
from 1/510 independent variables. We consider this last flaw fatal, and take the following
approach instead.

% Explain the problem (idea: comparing cosine similarities --> too much data, again)

Instead of comparing embedding pairs, we will compare embeddings
community by community.

At this stage, one might naively think that it would suffice to
calculate the distance between two embeddings index-wise, which is
equal to the Frobenius distance between \(L\) and \(S\):

@@

The problem with this approach is that even if several dimensions of
\(L\) and \(S\) are correlated, they will not coincide in the
\emph{representation}. That is, applying a simple rotation (orthogonal
transformation) on either matrix widely changes the \(||L-S||_F\)
correlation metric. Another way to consider the same problem is that,
applying such a rotation does not change the cosine similarities
considered in the previous section.

To make the metric independent of the representation (up to orthogonal
transformations), we compute the \emph{minimum} distance between
\(L_i\) and \(S_i\), for any orthogonal matrix \(\Omega\) applied to
\(L\):

@@

Here, the orthogonal matrix \(Ω\) gives a map from linguistic embeddings to social embeddings.
The problem of computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = n - \mathsf {Tr} (Σ)\] where the matrix \(Σ\) is obtained by the singular value
decomposition (SVD) \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). That is, each
singular value \(\sigma_i\) (the elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

@@

Now, to test if \(d(L,S)\) corresponds to a significant correlation,
it suffices to check if its value is significantly larger than it
value for random linguistic embeddings \(L\). Even though this is the
distribution of \(d(L,S)\) for random embeddings is difficult to
compute analytically, one can instead evaluate it using a Monte Carlo
method.

Doing so, we observed that \(d(L,S)\) exhibits a mean of
\(μ_d=431.39\) and a (Bessel's-corrected) standard deviation
\(s_d=2.90\) in their distance from the social embedding,
\(S\). % luckily it seems that sample stdev is often written with an $s$ to differentiate from population stdev

Thus if the real \(d(L,S)\) is below the mean by several standard
deviations, we can safely assume that there is statistically
significant correlation between \(L\) and \(S\). A 4-sigma difference
has less than one percent chance of occurring randomly. In our case:
we observe a difference between 61 and 68 standard deviations
(\cref{tab:embedding-correlations}). This definitely indicates a
significant correlation.  Furthermore, by coming back to the
definition of \(d(L,S)\), we know that, on average, the cosine
similarity between \(ΩL\) and \(S\) is \(510 - 232.18 / 510\).  It
further means that if we obtain a linguistic embedding \(L_k\) for a
new community $k$, we can estimate its social embedding by \(ΩL_k\),
and the cosine similarity with its true social embedding \(S_k\) is
expected to be \((431.39 - 232.18) / 510\) --- accounting for
over-fitting effects by taking the average distance rather than the
maximum. % JP: TODO is the number from the table. this should probably be rewritten
In sum, it is clear that the CCLM embeddings predict some aspect the social-network embeddings--- but far from all of it.

To finish, we also give a sense of \emph{how} the correlation is manifested overall, by
analysis of the two principal components of correlation in the
linguistic embeddings, \(U_0\) and \(U_1\). To do so we plot the
projection of each embedding along their first two principle components
which, together with the corresponding singular values, gives an idea
of how much and in what way they differ (see~\cref{fig:pca-aligned}).
